---
title: "Using isoForest"
author: "lucian xu"
date: "`r Sys.Date()`"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using 'isoForest'}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Overview

The {isoForest} package is a simple replication of the Isolation Forests
algorithm for outlier detection, and the {ranger} package is used to
truly construct the forests. In addition, the visualization of outliers
is also implemented to help better observe the prediction results.

Here's how isoForest was built:

1.  Use the ranger() function of the 'ranger' package to build the
    forest. Key parameter for Isolation Forests is:

    -   **sample_size: min(nrow(data), 256L)** *In the process of
        constructing a decision tree, when the sampling number is 256,
        the effect is the best.If the number of samples does not exceed
        256, take all samples.*
    -   **max_depth: ceiling(log2(sample_size))**
    -   **min.node.size: 1** *Every sample ultimately exists in an
        independent node.*
    -   **num.random.splits: 1** *Consider splitting based on only one
        feature each time.*
    -   **splitrule: "extratrees"** *The splitting point of the feature
        is random.*
    -   **replace: FALSE** *Perform sampling without replacement.*

2.  Obtain the depth information from each leaf node to the root node
    for every tree in the model.

3.  Use the original samples for prediction and obtain the predicted
    leaf node ID for each sample in each tree.

4.  Retrieve the depth of the predicted value for each sample in each
    tree and calculate the average depth across the entire forest for
    each sample.

5.  Calculate the anomaly score for each sample using the average depth
    across the entire forest.

## Installation

```{r install, eval=FALSE, include=FALSE}
# Development version
devtools::install_github("flystar233/isoForest")
```

## Usage

```{r usage, echo=TRUE}
library(isoForest)
result <- isoForest(iris)
head(result$scores)
```

## Setting Anomaly Thresholds

The package provides multiple methods to automatically determine anomaly thresholds:

```{r threshold, echo=TRUE}
# Train model
model <- isoForest(iris[1:4])

# Method 1: Contamination-based (most common)
result_contamination <- set_anomaly_threshold(
  model, 
  method = "contamination", 
  contamination = 0.05
)

# Method 2: MAD (robust to outliers)
result_mad <- set_anomaly_threshold(
  model, 
  method = "mad", 
  mad_multiplier = 3
)

# Method 3: KDE-weighted (density-aware)
result_kde <- set_anomaly_threshold(
  model, 
  method = "kde_weighted", 
  kde_multiplier = 3
)

# Method 4: MTT (statistical test, good for small samples)
result_mtt <- set_anomaly_threshold(
  model, 
  method = "mtt", 
  mtt_alpha = 0.05
)

# Compare results
cat("Contamination method:", sum(result_contamination$predictions$is_anomaly), "anomalies\n")
cat("MAD method:", sum(result_mad$predictions$is_anomaly), "anomalies\n")
cat("KDE-weighted method:", sum(result_kde$predictions$is_anomaly), "anomalies\n")
cat("MTT method:", sum(result_mtt$predictions$is_anomaly), "anomalies\n")
```

## Feature Contribution Analysis

Understand which features contribute most to a sample's anomaly score:

```{r contribution, echo=TRUE, fig.width=7, fig.height=5}
# Analyze feature contributions for specific samples
contributions <- feature_contribution(
  model, 
  sample_ids = c(42, 107), 
  data = iris[1:4],
  method = "path"
)

# Print results
print(contributions)
```

## Feature Distribution Visualization

### Visualizing Single Anomaly

When you want to understand why a specific sample is anomalous:

```{r boxplot_single, echo=TRUE, fig.width=8, fig.height=5}
# Calculate contributions
contributions <- feature_contribution(model, sample_ids = 42, data = iris[1:4])

# Plot boxplots showing where the anomaly falls
plot_feature_boxplot(contributions, iris[1:4], sample_id = 42, top_n = 4)
```

### Visualizing Multiple Anomalies

When you want to see where all detected anomalies fall in the distributions:

```{r boxplot_multiple, echo=TRUE, fig.width=8, fig.height=5}
# Detect anomalies
result <- set_anomaly_threshold(model, method = "mad", mad_multiplier = 3)
anomaly_ids <- which(result$predictions$is_anomaly)

# Visualize all anomalies at once (no contribution_obj needed)
plot_feature_boxplot(
  contribution_obj = NULL,
  data = iris[1:4],
  sample_id = anomaly_ids
)
```

### Faceted View for Multiple Features

```{r boxplot_faceted, echo=TRUE, fig.width=8, fig.height=6}
# Faceted view - better for viewing many features
plot_feature_boxplot_faceted(
  contribution_obj = NULL,
  data = iris[1:4],
  sample_id = anomaly_ids,
  top_n = 4,
  ncol = 2
)
```

**Key Features:**

- **Flexible Input**: Works with or without `contribution_obj`
- **Multiple Anomalies**: `sample_id` can be a vector to highlight many points
- **Auto-Optimization**: Point size adjusts automatically when >5 anomalies
- **Smart Labels**: Subtitle adapts based on number of anomalies
- **Jitter Effect**: All points use jitter to avoid overlap

## Basic Visualization

### Heatmap

```{r heatmap, echo=TRUE, fig.width=6, fig.height=4}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "heatmap")
```

### Scatter Plot

```{r scatter, echo=TRUE, fig.width=6, fig.height=4}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "scatter")
```

## Complete Workflow Example

Here's a complete example combining all features:

```{r complete_workflow, echo=TRUE, fig.width=8, fig.height=6}
# 1. Train model
model <- isoForest(iris[1:4])

# 2. Set threshold using robust method
result <- set_anomaly_threshold(model, method = "mad", mad_multiplier = 3)

# 3. Get anomaly IDs
anomaly_ids <- which(result$predictions$is_anomaly)
cat("Detected", length(anomaly_ids), "anomalies\n")

# 4. Analyze top anomaly
if (length(anomaly_ids) > 0) {
  top_anomaly <- anomaly_ids[1]
  contributions <- feature_contribution(model, sample_ids = top_anomaly, data = iris[1:4])
  
  cat("\nTop anomaly (Sample", top_anomaly, "):\n")
  print(contributions$sample_1$contributions)
  
  # 5. Visualize
  plot_feature_boxplot_faceted(contributions, iris[1:4], sample_id = top_anomaly)
}
```

## High-Dimensional Data Visualization

For high-dimensional data (more than 4 features), dimensionality reduction techniques like PCA and UMAP help visualize anomalies in 2D space. For large datasets, automatic smart sampling is applied to improve visualization speed while preserving all anomalies.

### PCA Visualization

```{r pca_viz, eval=FALSE}
# Quick PCA projection
plot_pca <- plot_anomalies_2d(
  model = model,
  data = iris[1:4],
  dim_reduction = "pca",
  method = "mtt"
)
print(plot_pca)
```

**Features:**
- Shows variance explained by PC1 and PC2
- Fast and stable
- Works with any dimensional data
- Automatic sampling for large datasets (anomalies always shown)

### UMAP Visualization

```{r umap_viz, eval=FALSE}
# Requires: install.packages("umap")

# UMAP projection (better for non-linear patterns)
plot_umap <- plot_anomalies_2d(
  model = model,
  data = iris[1:4],
  dim_reduction = "umap",
  method = "mtt",
  umap_n_neighbors = 15,
  umap_min_dist = 0.1
)
print(plot_umap)
```

**Features:**
- Preserves local and global structure
- Better reveals clusters
- Ideal for complex, non-linear relationships

### PCA vs UMAP Comparison

```{r comparison, eval=FALSE}
# Requires: install.packages(c("umap", "gridExtra"))

# Compare both methods side-by-side
comparison <- compare_dim_reduction(
  model = model,
  data = iris[1:4],
  method = "mtt"
)
```

### Smart Sampling for Large Datasets

For datasets with many points, the visualization functions automatically apply smart sampling to improve speed:

```{r smart_sampling, eval=FALSE}
# Default: anomalies represent 5% of displayed points
# If you have 100 anomalies, ~2000 total points will be shown
plot_anomalies_2d(model, large_data)

# Show fewer points (faster, anomalies = 10% of display)
plot_anomalies_2d(model, large_data, sample_rate = 0.10)

# Show more points (slower, anomalies = 2% of display)
plot_anomalies_2d(model, large_data, sample_rate = 0.02)

# Disable sampling (show all points)
plot_anomalies_2d(model, large_data, sample_rate = NULL)
```

**How it works:**
- All anomalies are **always shown**
- Normal points are randomly sampled
- Target: anomalies represent `sample_rate` proportion of displayed data
- Example: 50 anomalies with `sample_rate = 0.05` → show ~1000 total points

**When sampling is applied, the subtitle will show:**
```
Method: mtt | Anomalies: 50/5000 (1.0%) | Threshold: 0.635 | Showing: ~1000/5000 points
```

### When to Use Which Method?

| Scenario | Recommended Method |
|----------|-------------------|
| Quick exploration | PCA |
| Linear relationships | PCA |
| Non-linear patterns | UMAP |
| Cluster discovery | UMAP |
| Need interpretability | PCA |
| Complex high-dimensional data | UMAP |

### Recommended Workflow

1. **Start with PCA** for fast exploration
   ```r
   plot_anomalies_2d(model, data, dim_reduction = "pca")
   # Default sample_rate = 0.05 (anomalies = 5% of display)
   ```

2. **Try UMAP** if PCA doesn't separate anomalies well
   ```r
   plot_anomalies_2d(model, data, dim_reduction = "umap")
   ```

3. **Compare both** methods to see which works better
   ```r
   compare_dim_reduction(model, data)
   # Shows PCA and UMAP side-by-side with shared title
   ```

4. **Adjust sampling** if needed for speed or detail
   ```r
   # Faster visualization (fewer points)
   plot_anomalies_2d(model, data, sample_rate = 0.10)
   
   # More detail (more points)
   plot_anomalies_2d(model, data, sample_rate = 0.02)
   ```

5. **Validate with feature-level analysis**
   ```r
   plot_feature_boxplot_faceted(
     contribution_obj = NULL,
     data = data,
     sample_id = anomaly_ids
   )
   ```

This multi-level approach ensures you understand anomalies from both global (dimensionality reduction) and local (feature-level) perspectives. The smart sampling feature makes it practical to visualize large datasets without sacrificing insight into anomalies.
```
