---
title: "Using isoForest"
author: "lucian xu"
date: "`r Sys.Date()`"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using 'isoForest'}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Overview

The {isoForest} package is a simple replication of the Isolation Forests
algorithm for outlier detection, and the {ranger} package is used to
truly construct the forests. In addition, the visualization of outliers
is also implemented to help better observe the prediction results.

Here's how isoForest was built:

1.  Use the ranger() function of the 'ranger' package to build the
    forest. Key parameter for Isolation Forests is:

    -   **sample_size: min(nrow(data), 256L)** *In the process of
        constructing a decision tree, when the sampling number is 256,
        the effect is the best.If the number of samples does not exceed
        256, take all samples.*
    -   **max_depth: ceiling(log2(sample_size))**
    -   **min.node.size: 1** *Every sample ultimately exists in an
        independent node.*
    -   **num.random.splits: 1** *Consider splitting based on only one
        feature each time.*
    -   **splitrule: "extratrees"** *The splitting point of the feature
        is random.*
    -   **replace: FALSE** *Perform sampling without replacement.*

2.  Obtain the depth information from each leaf node to the root node
    for every tree in the model.

3.  Use the original samples for prediction and obtain the predicted
    leaf node ID for each sample in each tree.

4.  Retrieve the depth of the predicted value for each sample in each
    tree and calculate the average depth across the entire forest for
    each sample.

5.  Calculate the anomaly score for each sample using the average depth
    across the entire forest.

## Installation

```{r install, eval=FALSE, include=FALSE}
# Development version
devtools::install_github("flystar233/isoForest")
```

## Usage

```{r usage, echo=TRUE}
library(isoForest)
result <- isoForest(iris)
head(result$scores)
```

## Visualization

## heatmap

```{r heatmap, echo=TRUE}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "heatmap")
```

## scatter

```{r scatter, echo=TRUE}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "scatter")
```
