---
title: "Using isoForest"
author: "lucian xu"
date: "`r Sys.Date()`"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Using 'isoForest'}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.align = "center"
)
```

## Overview

The {isoForest} package is a simple replication of the Isolation Forests
algorithm for outlier detection, and the {ranger} package is used to
truly construct the forests. In addition, the visualization of outliers
is also implemented to help better observe the prediction results.

Here's how isoForest was built:

1.  Use the ranger() function of the 'ranger' package to build the
    forest. Key parameter for Isolation Forests is:

    -   **sample_size: min(nrow(data), 256L)** *In the process of
        constructing a decision tree, when the sampling number is 256,
        the effect is the best.If the number of samples does not exceed
        256, take all samples.*
    -   **max_depth: ceiling(log2(sample_size))**
    -   **min.node.size: 1** *Every sample ultimately exists in an
        independent node.*
    -   **num.random.splits: 1** *Consider splitting based on only one
        feature each time.*
    -   **splitrule: "extratrees"** *The splitting point of the feature
        is random.*
    -   **replace: FALSE** *Perform sampling without replacement.*

2.  Obtain the depth information from each leaf node to the root node
    for every tree in the model.

3.  Use the original samples for prediction and obtain the predicted
    leaf node ID for each sample in each tree.

4.  Retrieve the depth of the predicted value for each sample in each
    tree and calculate the average depth across the entire forest for
    each sample.

5.  Calculate the anomaly score for each sample using the average depth
    across the entire forest.

## Installation

```{r install, eval=FALSE, include=FALSE}
# Development version
devtools::install_github("flystar233/isoForest")
```

## Usage

```{r usage, echo=TRUE}
library(isoForest)
result <- isoForest(iris)
head(result$scores)
```

## Setting Anomaly Thresholds

The package provides multiple methods to automatically determine anomaly thresholds:

```{r threshold, echo=TRUE}
# Train model
model <- isoForest(iris[1:4])

# Method 1: Contamination-based (most common)
result_contamination <- set_anomaly_threshold(
  model, 
  method = "contamination", 
  contamination = 0.05
)

# Method 2: MAD (robust to outliers)
result_mad <- set_anomaly_threshold(
  model, 
  method = "mad", 
  mad_multiplier = 3
)

# Method 3: KDE-weighted (density-aware)
result_kde <- set_anomaly_threshold(
  model, 
  method = "kde_weighted", 
  kde_multiplier = 3
)

# Method 4: MTT (statistical test, good for small samples)
result_mtt <- set_anomaly_threshold(
  model, 
  method = "mtt", 
  mtt_alpha = 0.05
)

# Compare results
cat("Contamination method:", sum(result_contamination$predictions$is_anomaly), "anomalies\n")
cat("MAD method:", sum(result_mad$predictions$is_anomaly), "anomalies\n")
cat("KDE-weighted method:", sum(result_kde$predictions$is_anomaly), "anomalies\n")
cat("MTT method:", sum(result_mtt$predictions$is_anomaly), "anomalies\n")
```

## Feature Contribution Analysis

Understand which features contribute most to a sample's anomaly score:

```{r contribution, echo=TRUE, fig.width=7, fig.height=5}
# Analyze feature contributions for specific samples
contributions <- feature_contribution(
  model, 
  sample_ids = c(42, 107), 
  data = iris[1:4],
  method = "path"
)

# Print results
print(contributions)
```

## Feature Distribution Visualization

### Visualizing Single Anomaly

When you want to understand why a specific sample is anomalous:

```{r boxplot_single, echo=TRUE, fig.width=8, fig.height=5}
# Calculate contributions
contributions <- feature_contribution(model, sample_ids = 42, data = iris[1:4])

# Plot boxplots showing where the anomaly falls
plot_feature_boxplot(contributions, iris[1:4], sample_id = 42, top_n = 4)
```

### Visualizing Multiple Anomalies

When you want to see where all detected anomalies fall in the distributions:

```{r boxplot_multiple, echo=TRUE, fig.width=8, fig.height=5}
# Detect anomalies
result <- set_anomaly_threshold(model, method = "mad", mad_multiplier = 3)
anomaly_ids <- which(result$predictions$is_anomaly)

# Visualize all anomalies at once (no contribution_obj needed)
plot_feature_boxplot(
  contribution_obj = NULL,
  data = iris[1:4],
  sample_id = anomaly_ids
)
```

### Faceted View for Multiple Features

```{r boxplot_faceted, echo=TRUE, fig.width=8, fig.height=6}
# Faceted view - better for viewing many features
plot_feature_boxplot_faceted(
  contribution_obj = NULL,
  data = iris[1:4],
  sample_id = anomaly_ids,
  top_n = 4,
  ncol = 2
)
```

**Key Features:**

- **Flexible Input**: Works with or without `contribution_obj`
- **Multiple Anomalies**: `sample_id` can be a vector to highlight many points
- **Auto-Optimization**: Point size adjusts automatically when >5 anomalies
- **Smart Labels**: Subtitle adapts based on number of anomalies
- **Jitter Effect**: All points use jitter to avoid overlap

## Basic Visualization

### Heatmap

```{r heatmap, echo=TRUE, fig.width=6, fig.height=4}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "heatmap")
```

### Scatter Plot

```{r scatter, echo=TRUE, fig.width=6, fig.height=4}
result <- isoForest(iris[1:2])
anomaly_plot(result, iris[1:2], plot_type = "scatter")
```

## Complete Workflow Example

Here's a complete example combining all features:

```{r complete_workflow, echo=TRUE, fig.width=8, fig.height=6}
# 1. Train model
model <- isoForest(iris[1:4])

# 2. Set threshold using robust method
result <- set_anomaly_threshold(model, method = "mad", mad_multiplier = 3)

# 3. Get anomaly IDs
anomaly_ids <- which(result$predictions$is_anomaly)
cat("Detected", length(anomaly_ids), "anomalies\n")

# 4. Analyze top anomaly
if (length(anomaly_ids) > 0) {
  top_anomaly <- anomaly_ids[1]
  contributions <- feature_contribution(model, sample_ids = top_anomaly, data = iris[1:4])
  
  cat("\nTop anomaly (Sample", top_anomaly, "):\n")
  print(contributions$sample_1$contributions)
  
  # 5. Visualize
  plot_feature_boxplot_faceted(contributions, iris[1:4], sample_id = top_anomaly)
}
```
